Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-08-19 paper/paper_with_refs.tex

\bibliography{references}

>>>>

\begin{thebibliography}{47}

\bibitem[{Bedford {\em et~al.\/}}(2020){Bedford, Greninger, Roychoudhury,
  Starita, Famulare, Huang, Nalla, Pepper, Reinhardt, Xie, {\em
  et~al.\/}}]{bedford2020cryptic}
{Bedford, T., A.~L. Greninger, P.~Roychoudhury, L.~M. Starita, M.~Famulare,
  {\em et~al.\/}}, 2020 Cryptic transmission of {SARS-CoV-2} in washington
  state. Science {\bf 370}: 571--575.

\bibitem[{Boni {\em et~al.\/}}(2020){Boni, Lemey, Jiang, Lam, Perry, Castoe,
  Rambaut, {\rm and} Robertson}]{boni2020evolutionary}
{Boni, M.~F., P.~Lemey, X.~Jiang, T.~T.-Y. Lam, B.~W. Perry, {\em et~al.\/}},
  2020 {Evolutionary origins of the SARS-CoV-2 sarbecovirus lineage responsible
  for the COVID-19 pandemic}. Nature Microbiology {\bf 5}: 1408--1417.

\bibitem[{Chan {\em et~al.\/}}(2020){Chan, Yuan, Kok, To, Chu, Yang, Xing, Liu,
  Yip, Poon, {\em et~al.\/}}]{chan2020familial}
{Chan, J. F.-W., S.~Yuan, K.-H. Kok, K.~K.-W. To, H.~Chu, {\em et~al.\/}}, 2020
  A familial cluster of pneumonia associated with the 2019 novel coronavirus
  indicating person-to-person transmission: a study of a family cluster. Lancet
  {\bf 395}: 514--523.

\bibitem[{Chen {\em et~al.\/}}(2020){Chen, Zhou, Dong, Qu, Gong, Han, Qiu,
  Wang, Liu, Wei, {\em et~al.\/}}]{chen2020epidemiological}
{Chen, N., M.~Zhou, X.~Dong, J.~Qu, F.~Gong, {\em et~al.\/}}, 2020
  Epidemiological and clinical characteristics of 99 cases of 2019 novel
  coronavirus pneumonia in {Wuhan, China}: a descriptive study. Lancet {\bf
  395}: 507--513.

\bibitem[{Chen {\em et~al.\/}}(2018){Chen, Zhou, Chen, {\rm and}
  Gu}]{chen2018fastp}
{Chen, S., Y.~Zhou, Y.~Chen, {\rm and} J.~Gu}, 2018 fastp: an ultra-fast
  all-in-one fastq preprocessor. Bioinformatics {\bf 34}: i884--i890.

\bibitem[{{China CDC}}(2020)]{chinacdc2020}
{{China CDC}}, 2020 {Memo to the Offices of the Chinese Center for Disease
  Control and Prevention}.
  \url{https://www.documentcloud.org/documents/7340336-China-CDC-Sup-Regs.html}.

\bibitem[{Cohen}(2020)]{cohen2020wuhan}
{Cohen, J.}, 2020 Wuhan seafood market may not be source of novel virus
  spreading globally. Science {\bf 10}: 10.1126/science.abb0611.

\bibitem[{Deng {\em et~al.\/}}(2020){Deng, Gu, Federman, Du~Plessis, Pybus,
  Faria, Wang, Yu, Bushnell, Pan, {\em et~al.\/}}]{deng2020genomic}
{Deng, X., W.~Gu, S.~Federman, L.~Du~Plessis, O.~G. Pybus, {\em et~al.\/}},
  2020 Genomic surveillance reveals multiple introductions of {SARS-CoV-2 into
  Northern California}. Science {\bf 369}: 582--587.

\bibitem[{Farkas {\em et~al.\/}}(2020){Farkas, Fuentes-Villalobos, Garrido,
  Haigh, {\rm and} Barr{\'\i}a}]{farkas2020insights}
{Farkas, C., F.~Fuentes-Villalobos, J.~L. Garrido, J.~Haigh, {\rm and} M.~I.
  Barr{\'\i}a}, 2020 Insights on early mutational events in {SARS-CoV-2} virus
  reveal founder effects across geographical regions. PeerJ {\bf 8}: e9255.

\bibitem[{Fauver {\em et~al.\/}}(2020){Fauver, Petrone, Hodcroft, Shioda,
  Ehrlich, Watts, Vogels, Brito, Alpert, Muyombwe, {\em
  et~al.\/}}]{fauver2020coast}
{Fauver, J.~R., M.~E. Petrone, E.~B. Hodcroft, K.~Shioda, H.~Y. Ehrlich, {\em
  et~al.\/}}, 2020 {Coast-to-coast spread of SARS-CoV-2 during the early
  epidemic in the United States}. Cell {\bf 181}: 990--996.

\bibitem[{Forster {\em et~al.\/}}(2020){Forster, Forster, Renfrew, {\rm and}
  Forster}]{forster2020phylogenetic}
{Forster, P., L.~Forster, C.~Renfrew, {\rm and} M.~Forster}, 2020 Phylogenetic
  network analysis of {SARS-CoV-2} genomes. Proceedings of the National Academy
  of Sciences {\bf 117}: 9241--9243.

\bibitem[{Garry}(2021)]{garry2021early}
{Garry, R.~F.}, 2021 Early appearance of two distinct genomic lineages of
  {SARS-CoV-2} in different wuhan wildlife markets suggests {SARS-CoV-2} has a
  natural origin.
  \url{https://virological.org/t/early-appearance-of-two-distinct-genomic-lineages-of-sars-cov-2-in-different-wuhan-wildlife-markets-suggests-sars-cov-2-has-a-natural-origin/691}.

\bibitem[{{Global Times}}(2020)]{GlobalTimes2020}
{{Global Times}}, 2020 Wuhan's huanan seafood market a victim of {COVID-19}:
  {CDC} director. \url{https://www.globaltimes.cn/content/1189506.shtml},
  archived at
  \url{https://web.archive.org/web/20200528062530/https://www.globaltimes.cn/content/1189506.shtml}.

\bibitem[{{Health Times}}(2020)]{healthtimes2020}
{{Health Times}}, 2020 Experts judge the source of the new coronavirus:
  {December} 8 last year may not be the earliest time of onset.
  \url{https://www.guancha.cn/politics/2020_02_27_538822.shtml}, archived and
  detailed at
  \url{https://docs.google.com/document/d/e/2PACX-1vTQxG822DtqP7IZSjLj751Mrm8Ev7leksXfjBLsA9KJ0_tbGV6YJAAjuijPnwz_YmUQGY1PZUl5LcCI/pub}.

\bibitem[{Huang {\em et~al.\/}}(2020){Huang, Wang, Li, Ren, Zhao, Hu, Zhang,
  Fan, Xu, Gu, {\em et~al.\/}}]{huang2020clinical}
{Huang, C., Y.~Wang, X.~Li, L.~Ren, J.~Zhao, {\em et~al.\/}}, 2020 Clinical
  features of patients infected with 2019 novel coronavirus in {Wuhan, China}.
  Lancet {\bf 395}: 497--506.

\bibitem[{Huerta-Cepas {\em et~al.\/}}(2016){Huerta-Cepas, Serra, {\rm and}
  Bork}]{huerta2016ete}
{Huerta-Cepas, J., F.~Serra, {\rm and} P.~Bork}, 2016 Ete 3: reconstruction,
  analysis, and visualization of phylogenomic data. Molecular biology and
  evolution {\bf 33}: 1635--1638.

\bibitem[{Kang {\em et~al.\/}}(2020{a}){Kang, Cheng, {\rm and}
  McNeil}]{Kang2020}
{Kang, D., M.~Cheng, {\rm and} S.~McNeil}, 2020{a} China clamps down in hidden
  hunt for coronavirus origins.
  \url{https://apnews.com/article/united-nations-coronavirus-pandemic-china-only-on-ap-bats-24fbadc58cee3a40bca2ddf7a14d2955},
  The actual China State Council order described in the article is at
  \url{https://www.documentcloud.org/documents/7340337-State-Research-regulations.html}.

\bibitem[{Kang {\em et~al.\/}}(2020{b}){Kang, Wu, Ma, He, Lu, Liu, Li, Mei,
  Ruan, Lin, Zou, Ke, Zhong, Zhang, Chen, Liu, Zhu, Xiao, Yu, Hu, Zeng, Li,
  Liao, Tang, Xiao, Wang, Song, Zhuang, Liang, Zeng, He, Lin, Deng, {\rm and}
  Song}]{kang2020evidence}
{Kang, M., J.~Wu, W.~Ma, J.~He, J.~Lu, {\em et~al.\/}}, 2020{b} Evidence and
  characteristics of human-to-human transmission of sars-cov-2. medRxiv .

\bibitem[{Katoh {\rm and} Standley}(2013)]{katoh2013mafft}
{Katoh, K. {\rm and} D.~M. Standley}, 2013 Mafft multiple sequence alignment
  software version 7: improvements in performance and usability. Molecular
  biology and evolution {\bf 30}: 772--780.

\bibitem[{Kumar {\em et~al.\/}}(2021){Kumar, Tao, Weaver, Sanderford,
  Caraballo-Ortiz, Sharma, Pond, {\rm and} Miura}]{kumar2021evolutionary}
{Kumar, S., Q.~Tao, S.~Weaver, M.~Sanderford, M.~A. Caraballo-Ortiz, {\em
  et~al.\/}}, 2021 An evolutionary portrait of the progenitor {SARS-CoV-2} and
  its dominant offshoots in {COVID-19} pandemic. Molecular Biology and
  Evolution msab118.

\bibitem[{Li}(2018)]{li2018minimap2}
{Li, H.}, 2018 Minimap2: pairwise alignment for nucleotide sequences.
  Bioinformatics {\bf 34}: 3094--3100.

\bibitem[{Li {\em et~al.\/}}(2021){Li, Wang, Ma, Li, Yang, Shi, {\rm and}
  Duan}]{li2021novel}
{Li, L., J.~Wang, X.~Ma, J.~Li, X.~Yang, {\em et~al.\/}}, 2021 A novel
  sars-cov-2 related virus with complex recombination isolated from bats in
  yunnan province, china. bioRxiv .

\bibitem[{Lytras {\em et~al.\/}}(2021){Lytras, Hughes, Martin, de~Klerk,
  Lourens, Kosakovsky~Pond, Xia, Jiang, {\rm and}
  Robertson}]{lytras2021exploring}
{Lytras, S., J.~Hughes, D.~Martin, A.~de~Klerk, R.~Lourens, {\em et~al.\/}},
  2021 {Exploring the natural origins of SARS-CoV-2 in the light of
  recombination}. bioRxiv .

\bibitem[{Ma}(2020)]{ma2020SCMP}
{Ma, J.}, 2020 {Coronavirus: China’s first confirmed Covid-19 case traced
  back to November 17}. South China Morning Post,
  \url{https://www.scmp.com/news/china/society/article/3074991/coronavirus-chinas-first-confirmed-covid-19-case-traced-back},
  archived at
  \url{https://web.archive.org/web/20200315011702/https://www.scmp.com/news/china/society/article/3074991/coronavirus-chinas-first-confirmed-covid-19-case-traced-back}.

\bibitem[{Minh {\em et~al.\/}}(2020){Minh, Schmidt, Chernomor, Schrempf,
  Woodhams, Von~Haeseler, {\rm and} Lanfear}]{minh2020iq}
{Minh, B.~Q., H.~A. Schmidt, O.~Chernomor, D.~Schrempf, M.~D. Woodhams, {\em
  et~al.\/}}, 2020 {IQ-TREE} 2: New models and efficient methods for
  phylogenetic inference in the genomic era. Molecular Biology and Evolution
  {\bf 37}: 1530--1534.

\bibitem[{M{\"o}lder {\em et~al.\/}}(2021){M{\"o}lder, Jablonski, Letcher,
  Hall, Tomkins-Tinch, Sochat, Forster, Lee, Twardziok, Kanitz, {\em
  et~al.\/}}]{molder2021sustainable}
{M{\"o}lder, F., K.~P. Jablonski, B.~Letcher, M.~B. Hall, C.~H. Tomkins-Tinch,
  {\em et~al.\/}}, 2021 Sustainable data analysis with snakemake. F1000Research
  {\bf 10}: 33.

\bibitem[{Morel {\em et~al.\/}}(2021){Morel, Barbera, Czech, Bettisworth,
  H{\"u}bner, Lutteropp, Serdari, Kostaki, Mamais, Kozlov, {\em
  et~al.\/}}]{morel2021phylogenetic}
{Morel, B., P.~Barbera, L.~Czech, B.~Bettisworth, L.~H{\"u}bner, {\em
  et~al.\/}}, 2021 Phylogenetic analysis of {SARS-CoV-2} data is difficult.
  Molecular Biology and Evolution {\bf 38}: 1777--1791.

\bibitem[{Pekar {\em et~al.\/}}(2021){Pekar, Worobey, Moshiri, Scheffler, {\rm
  and} Wertheim}]{pekar2021timing}
{Pekar, J., M.~Worobey, N.~Moshiri, K.~Scheffler, {\rm and} J.~O. Wertheim},
  2021 Timing the {SARS-CoV-2} index case in hubei province. Science {\bf 372}:
  412--417.

\bibitem[{Pingui}(2020)]{pinghui2020SCMP}
{Pingui, Z.}, 2020 China confirms unauthorised labs were told to destroy early
  coronavirus samples. South China Morning Post,
  \url{https://www.scmp.com/news/china/society/article/3084635/china-confirms-unauthorised-labs-were-told-destroy-early},
  archived at
  \url{https://web.archive.org/web/20210103124552/https://www.scmp.com/news/china/society/article/3084635/china-confirms-unauthorised-labs-were-told-destroy-early}.

\bibitem[{Pipes {\em et~al.\/}}(2021){Pipes, Wang, Huelsenbeck, {\rm and}
  Nielsen}]{pipes2021assessing}
{Pipes, L., H.~Wang, J.~P. Huelsenbeck, {\rm and} R.~Nielsen}, 2021 Assessing
  uncertainty in the rooting of the {SARS-CoV-2} phylogeny. Molecular Biology
  and Evolution {\bf 38}: 1537--1543.

\bibitem[{ProMED}(2019)]{ProMED2109}
{ProMED}, 2019 Undiagnosed pneumonia---{China (Hubei)}: Request for
  information. \url{https://promedmail.org/promed-post/?id=6864153}.

\bibitem[{Rambaut {\em et~al.\/}}(2020){Rambaut, Holmes, O’Toole, Hill,
  McCrone, Ruis, du~Plessis, {\rm and} Pybus}]{rambaut2020dynamic}
{Rambaut, A., E.~C. Holmes, {\'A}.~O’Toole, V.~Hill, J.~T. McCrone, {\em
  et~al.\/}}, 2020 A dynamic nomenclature proposal for {SARS-CoV-2} lineages to
  assist genomic epidemiology. Nature Microbiology {\bf 5}: 1403--1407.

\bibitem[{Shu {\rm and} McCauley}(2017)]{shu2017gisaid}
{Shu, Y. {\rm and} J.~McCauley}, 2017 {GISAID}: Global initiative on sharing
  all influenza data--from vision to reality. Eurosurveillance {\bf 22}: 30494.

\bibitem[{SRA}(2021)]{SRA_deletion}
{SRA}, 2021 {SRA} data updates.
  \url{https://www.ncbi.nlm.nih.gov/sra/docs/submitupdate/#how-do-i-withdraw-sra-data},
  Last accessed June 2, 2021; a copy of the page is digitally archived on the
  Wayback Machine at
  \url{https://web.archive.org/web/20210505002105/https://www.ncbi.nlm.nih.gov/sra/docs/submitupdate/}.

\bibitem[{Turakhia {\em et~al.\/}}(2020){Turakhia, De~Maio, Thornlow, Gozashti,
  Lanfear, Walker, Hinrichs, Fernandes, Borges, Slodkowicz, {\em
  et~al.\/}}]{turakhia2020stability}
{Turakhia, Y., N.~De~Maio, B.~Thornlow, L.~Gozashti, R.~Lanfear, {\em
  et~al.\/}}, 2020 Stability of {SARS-CoV-2} phylogenies. PLoS Genetics {\bf
  16}: e1009175.

\bibitem[{VanderPlas {\em et~al.\/}}(2018){VanderPlas, Granger, Heer, Moritz,
  Wongsuphasawat, Satyanarayan, Lees, Timofeev, Welsh, {\rm and}
  Sievert}]{vanderplas2018altair}
{VanderPlas, J., B.~E. Granger, J.~Heer, D.~Moritz, K.~Wongsuphasawat, {\em
  et~al.\/}}, 2018 Altair: interactive statistical visualizations for {Python}.
  Journal of Open Source Software {\bf 3}: 1057.

\bibitem[{Wang {\em et~al.\/}}(2020{a}){Wang, Fu, Hu, Tong, Liu, Gu, Liu,
  Jiang, Shen, Zhao, Men, Deng, Yu, Li, {\rm and} Liu}]{wang2020medRxiv}
{Wang, M., A.~Fu, B.~Hu, Y.~Tong, R.~Liu, {\em et~al.\/}}, 2020{a} Nanopore
  target sequencing for accurate and comprehensive detection of {SARS-CoV-2}
  and other respiratory viruses. medRxiv {\bf 10.1101/2020.03.04.20029538}.

\bibitem[{Wang {\em et~al.\/}}(2020{b}){Wang, Fu, Hu, Tong, Liu, Liu, Gu,
  Xiang, Liu, Jiang, Shen, Zhao, Men, Deng, Yu, Wei, Li, {\rm and}
  Liu}]{wang2020small}
{Wang, M., A.~Fu, B.~Hu, Y.~Tong, R.~Liu, {\em et~al.\/}}, 2020{b} Nanopore
  targeted sequencing for the accurate and comprehensive detection of
  {SARS-CoV-2} and other respiratory viruses. Small {\bf 16}: 2002169.

\bibitem[{Wang}(2021{a})]{wang2021exclusive}
{Wang, Z.}, 2021{a} {Exclusive: Chinese researcher, at the center of COVID data
  withdrawal controversy, speaks up}.
  \url{https://beijingchannel.substack.com/p/exclusive-chinese-researcher-at-the}.

\bibitem[{Wang}(2021{b})]{wang2021why}
{Wang, Z.}, 2021{b} {Why did Wuhan University researchers delete COVID-19 data
  at NIH?}
  \url{https://pekingnology.substack.com/p/why-did-wuhan-university-researchers}.

\bibitem[{{WHO}}(2021)]{WHO2021origins}
{{WHO}}, 2021 {WHO-convened global study of origins of SARS-CoV-2: China Part}.
  \url{https://www.who.int/publications/i/item/who-convened-global-study-of-origins-of-sars-cov-2-china-part}.

\bibitem[{Worobey {\em et~al.\/}}(2020){Worobey, Pekar, Larsen, Nelson, Hill,
  Joy, Rambaut, Suchard, Wertheim, {\rm and} Lemey}]{worobey2020emergence}
{Worobey, M., J.~Pekar, B.~B. Larsen, M.~I. Nelson, V.~Hill, {\em et~al.\/}},
  2020 {The emergence of SARS-CoV-2 in Europe and North America}. Science {\bf
  370}: 564--570.

\bibitem[{Yang {\em et~al.\/}}(2020){Yang, Lu, Liu, Wang, Zhang, Jalali, Dean,
  Longini, Halloran, Xu, Zhang, Wang, Liu, {\rm and}
  Fang}]{yang2020epidemiological}
{Yang, Y., Q.-B. Lu, M.-J. Liu, Y.-X. Wang, A.-R. Zhang, {\em et~al.\/}}, 2020
  Epidemiological and clinical features of the 2019 novel coronavirus outbreak
  in china. medRxiv {\bf 10.1101/2020.02.10.20021675}.

\bibitem[{Zhou {\em et~al.\/}}(2020{a}){Zhou, Chen, Hu, Li, Song, Liu, Wang,
  Liu, Yang, Holmes, {\em et~al.\/}}]{zhou2020novel}
{Zhou, H., X.~Chen, T.~Hu, J.~Li, H.~Song, {\em et~al.\/}}, 2020{a} A novel bat
  coronavirus closely related to sars-cov-2 contains natural insertions at the
  s1/s2 cleavage site of the spike protein. Current Biology {\bf 30}:
  2196--2203.

\bibitem[{Zhou {\em et~al.\/}}(2021){Zhou, Ji, Chen, Bi, Li, Wang, Hu, Song,
  Zhao, Chen, {\em et~al.\/}}]{zhou2021identification}
{Zhou, H., J.~Ji, X.~Chen, Y.~Bi, J.~Li, {\em et~al.\/}}, 2021 Identification
  of novel bat coronaviruses sheds light on the evolutionary origins of
  sars-cov-2 and related viruses. Cell .

\bibitem[{Zhou {\em et~al.\/}}(2020{b}){Zhou, Yang, Wang, Hu, Zhang, Zhang, Si,
  Zhu, Li, Huang, {\em et~al.\/}}]{zhou2020pneumonia}
{Zhou, P., X.-L. Yang, X.-G. Wang, B.~Hu, L.~Zhang, {\em et~al.\/}}, 2020{b} A
  pneumonia outbreak associated with a new coronavirus of probable bat origin.
  Nature {\bf 579}: 270--273.

\bibitem[{Zhu {\em et~al.\/}}(2020){Zhu, Zhang, Wang, Li, Yang, Song, Zhao,
  Huang, Shi, Lu, {\em et~al.\/}}]{zhu2020brief}
{Zhu, N., D.~Zhang, W.~Wang, X.~Li, B.~Yang, {\em et~al.\/}}, 2020 Brief
  report: A novel coronavirus from patients with pneumonia in china, 2019. The
  New England Journal of Medicine {\bf 382}: 727.

\end{thebibliography}
